For Finitary Induction-Induction, \\ Induction is Enough

### 2020-07-08 paper/paper.tex
newest lipics style and some small fixes in the paper

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no.\ ED18-1-2019-0030 (Application-specific highly reliable IT
solutions), by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no.\ BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@gmail.com}{https://orcid.org/0000-0002-9299-641X}{this author was supported by the CoqHoTT ERC Grant 637339.}

@@

\Copyright{A. Kaposi, A. Kov{\'a}cs and A. Lafont}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\category{}%optional, e.g. invited paper

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive
  types in type theory. They allow the mutual definition of types with
  multiple sorts where later sorts can be indexed by previous ones. An
  example is the Chapman-style syntax of type theory with conversion
  relations for each sort where e.g.\ the sort of types is indexed by
  contexts. In this paper we show that if a model of extensional type
  theory (ETT) supports indexed W-types, then it supports finitely
  branching IITs. We use a small internal type theory called the
  theory of signatures to specify signatures for IITs. We show that if
  a model of ETT supports the syntax for the theory of signatures,
  then it supports all IITs. We construct this syntax from indexed
  W-types using preterms and typing relations and prove its initiality
  following Streicher. The construction of the syntax and its
  initiality proof were formalised in Agda.
\end{abstract}

@@

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\GG:\Con)\ra Con^D\,\GG \\
  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra Ty^D\,(\mathsf{elimCon}\,\GG)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = {ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = U^D\,(\mathsf{elimCon}\,\GG) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = El^D\,(\mathsf{elimCon}\,\GG)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and it is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} presented a reduction in
extensional type theory, however, this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak, e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.

@@

In this paper we show that any model of extensional type theory with
indexed W-types supports the theory of IIT signatures, and as a
consequence all IITs. The difficulty in this construction is that the
theory of IIT signatures is itself a QIIT, it is both
inductive-inductive and has equality constructors. However, it can be
seen as the well-typed syntax of a small type theory without any
computation rules. Hence we can represent the syntax of normal forms
without quotienting. We construct this well-typed normal syntax using
preterms and typing relations from indexed W-types. Finally, we prove
the elimination principle in the style of the initiality proof of
Streicher \cite{streichersemantics}.

@@

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. In \Cref{thm:ToSToIITs}, we show that if a
  model of ETT supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs in the previously
  given. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
\end{itemize}
We list further work in \Cref{sec:conclusions}.

>>>>

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no.\ ED\_18-1-2019-0030, by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no.\ BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@gmail.com}{https://orcid.org/0000-0002-9299-641X}{this author was supported by the CoqHoTT ERC Grant 637339.}

@@

\Copyright{Ambrus Kaposi, Andr{\'a}s Kov{\'a}cs and Ambroise Lafont}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

%\category{}%optional, e.g. invited paper

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Marc Bezem and Assia Mahboubi}
\EventNoEds{2}
\EventLongTitle{25th International Conference on Types for Proofs and Programs (TYPES 2019)}
\EventShortTitle{TYPES 2019}
\EventAcronym{TYPES}
\EventYear{2019}
\EventDate{June 11--14, 2019}
\EventLocation{Oslo, Norway}
\EventLogo{}
\SeriesVolume{175}
\ArticleNo{6}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive
  types in type theory. They allow the mutual definition of types with
  multiple sorts where later sorts can be indexed by previous ones. An
  example is the Chapman-style syntax of type theory with conversion
  relations for each sort where e.g.\ the sort of types is indexed by
  contexts. In this paper we show that if a model of extensional type
  theory (ETT) supports indexed W-types, then it supports finitely
  branching IITs. We use a small internal type theory called the
  theory of signatures to specify IITs. We show that if a model of ETT
  supports the syntax for the theory of signatures, then it supports
  all IITs. We construct this syntax from indexed W-types using
  preterms and typing relations and prove its initiality following
  Streicher. The construction of the syntax and its initiality proof
  were formalised in Agda.
\end{abstract}

@@

Given the above motives and methods the elimination principle provides
two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\GG:\Con)\ra Con^D\,\GG \\
  & \mathsf{elimTy} && : (A:\Ty\,\GG)\ra Ty^D\,(\mathsf{elimCon}\,\GG)\,A
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = {empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\GG\,A) && = {ext}^D\,(\mathsf{elimCon}\,\GG)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\GG) && = U^D\,(\mathsf{elimCon}\,\GG) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\GG) && = El^D\,(\mathsf{elimCon}\,\GG)
\end{alignat*}
The functions $\mathsf{elimCon}$ and $\mathsf{elimTy}$ are an example
of a \emph{recursive-recursive} definition (using nomenclature from
\cite{forsberg-phd}). This means two mutually defined functions where
the type of the second function depends on the first function. The
proof assistant Agda \cite{norell07thesis} allows defining such
functions (even from non-IITs) and is currently the only proof
assistant supporting IITs\footnote{An experimental version of Coq with
  IITs is also available on GitHub.}.

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} presented a reduction in
extensional type theory, however, this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives, methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is not capable of defining standard
(metacircular) interpretation \cite{ttintt} of our small syntax. Using
pattern matching notation, this interpretation is the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.

@@

In this paper we show that any model of extensional type theory with
indexed W-types supports the theory of IIT signatures, and as a
consequence all IITs. The difficulty in this construction is that the
theory of IIT signatures is itself a QIIT, it is both
inductive-inductive and has equality constructors. However, it can be
seen as the well-typed syntax of a small type theory without any
computation rules. Hence we can represent the syntax of normal forms
without quotienting. We construct this well-typed normal syntax using
preterms and typing relations from indexed W-types. Finally, we prove
the elimination principle in the style of the initiality proof of
Streicher.

Streicher \cite{streichersemantics} constructs the syntactic model of
type theory using well-typed preterms and then shows initiality of
this model by (1) defining a partial map to any other model by
induction on preterms and (2) showing that whenever this partial
function receives a well-typed preterm on its input it actually gives
an output. Instead of defining a partial function, we define the graph
of the same function as a relation and then show that it is functional
as a second step. This can be seen as an indexed variant of the
construction using partial functions.

@@

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. In \Cref{thm:ToSToIITs}, we show that if a
  model of ETT supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
\end{itemize}
We list further work in \Cref{sec:conclusions}.
